Multimodal Spatial Availability: A Singly-Constrained Measure of Competitive Accessibility by Multiple Modes

### 2023-08-24 Journal-manuscript/Journal-manuscript.Rmd
Submitted this version to journal 1

---
title: "Competition for Access to Opportunities between Modes: Multimodal Spatial Availability"
#runningheader: "Multimodal Spatial Availability: A Singly-Constrained Measure of Competitive Accessibility by Multiple Modes"
author:
  - name: Anastasia Soukhov
    note: Corresponding Author
    position: PhD Candidate
    email: soukhoa@mcmaster.ca
    affiliation: Department of Earth, Environment and Society, McMaster University, Canada
  - name: Javier Tarriño-Ortiz
    position: PhD Graduate
    email: javier.tarrino.ortiz@upm.es
    affiliation: Universidad Politécnica de Madrid, Spain
  - name: Julio A. Soria-Lara
    position: Associate Professor
    email: julio.soria-lara@upm.es
    affiliation: Urban and Regional Planning, Universidad Politécnica de Madrid, Spain
  - name: Antonio Páez
    position: Associate Professor
    email: paezha@mcmaster.ca
    affiliation: Department of Earth, Environment and Society, McMaster University, Canada
abstract: |
  An increasing number of studies within the domain of transportation planning are concerned with the inequities in accessibility to opportunities. A dimension of these inequities arises from differences in access by mode type (e.g., commuting using a car as opposed to transit). However, methods implemented in current accessibility literature are lacking within the context of multimodal analysis. This paper presents an extension of spatial availability, a singly-constrained competitive accessibility measure, for the context of multimodal accessibility analysis. We first illustrate the features of spatial availability that lend itself to multimodal analysis. We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes. In summary, spatial availability can be used to create and interpret multimodal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape.

keywords: ["Multimodal", "Accessibility", "Equity", "Policy scenarios", "Low emission zones"]
bibliography: mybibfile.bib
biblio-style: unsrtnat
classoption:
#- numbered #  When numbered option is activated, lines are numbered.
output: 
  #word_document
  rticles::trb_article
wordcount: 7195
---
```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = '', 
  out.width = "1\\linewidth")

@@

library(patchwork) # The Composer of Plots
library(sf) # Simple Features for R
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggforce) # Accelerating 'ggplot2'
library(geomtextpath) # Curved Text in 'ggplot2'
library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
library(ggspatial) # Spatial Data Framework for ggplot2
library(flextable) # Functions for Tabular Reporting
library(scales) # Scale Functions for Visualization
library(TAM)
library(officer)
library(ggnewscale)
library(bookdown)
```

@@

# od_lines coded as segments
od_lines <- data.frame(x = c(2.5, 2.5, 2.5, 2.5, 2.5, 2.5, 6.5, 6.5, 6.5),
                       y = c(8.0, 8.0, 8.0, 2.5, 2.5, 2.5, 4.5, 4.5, 4.5),
                       xend = c(0.5, 0.5, 6.5, 0.5, 0.5, 6.5, 0.5, 0.5, 6.5),
                       yend = c(7.5, 2.5, 2.5, 7.5, 2.5, 2.5, 7.5, 2.5, 2.5),
                       group = c("A", "A", "A", "B", "B", "B", "C", "C", "C"),
                       time = c("15 min (x)\n10 min (z)", "30 min (x)\n25 min (z)", "100 min (x)\n80 min (z)",
                                "30 min (x)\n25 min (z)", "15 min (x)\n10 min (z)", "100 min (x)\n80 min (z)",
                                "100 min (x)\n80 min (z)", "100 min (x)\n80 min (z)", "15 min (x)\n10 min (z)"))

@@

```{r, creating-toy-example-plot, eval=FALSE}
toy_example_plot <- ggplot() + 
  # Plot centers
  geom_circle(data = centers,
             aes(x0 = x, 
                 y0 = y,
                 r = radius),
             color = "grey40") +
  annotate(geom = "label", 
           x=c(0.5, 0.6, 7),
           y=c(9, 1, 5.2), 
           label = c("Suburban", "Urban", "Satellite"), 
           size=4, 
           label.size=NA) +
  # Connect origins and destinations with curves
  geom_curve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group),
             curvature = 0.20,
             linewidth = 1,
             color = "black") +
  # Text on curves
  geom_textcurve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group,
                   label = time),
              size = 3,
             curvature = 0.20,
             color = "black") +
  # Plot origins and destinations
  geom_point(data = od,
            aes(size = size,
                x = x, 
                y = y, 
                color = type,
                shape = type)) +
  scale_size(range = c(4, 10)) +
  # Label origins and destinations
  geom_text(data = od, 
            aes(x, 
                y, 
                label=id), 
            size=3) +
  # Create table
  annotate(geom = "table",
           x = 8, 
           y = 9, 
           label = list(od_table),
           fill = "white",
           size=3) +
  # Theme the plot
  coord_fixed() + 
  theme_void() + guides(shape = "none", size = "none", color = "none", linetype ="none")

@@

sf::sf_use_s2(FALSE) #this terns off the spherical geometry component in the st_union() function - this assumes the sf is just planar. Thats' fine for our purposes
Madrid_muni_border <- zn1259_mad %>%
  st_union()
```
```{r merging-borders-for-plotting}
Madrid_muni_border <- Madrid_muni_border %>% st_as_sf() %>% transmute(Border = "Madrid municipal border") %>% rename( "geometry" = "x")

borders <- rbind(LEZ_centro %>% transmute(Border = "LEZ centro"), 
                 LEZ_M30 %>% transmute(Border = "LEZ M-30"),
                 Madrid_muni_border)
```

@@

```{r}
#Plot v (spatial availability per capita) by origin:
SA_im_v_zn208_plot <- ggplot() +
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=NA,
          fill="grey30") +
  geom_sf(data = SA_im_zn208,
          aes(fill = v_im),
          color = NA) +
  scale_fill_gradient2(low="red", high="green", mid = "white", midpoint = 1,
                       name = expression(v["i"]^{m}),
                       limits = c(0.001106867, 2.086736720),
                       breaks = c( 0.12, 0.31, 0.80, 1.11, 2.086736720),
                       labels = c(">0.12 (20% prob.)", "0.31 (40% prob.)", "0.80 (60% prob.)", "1.11 (80% prob.)", "2.09 (max.)"))+
    geom_sf(data = borders,
          aes(colour=Border),
          size = 0.5,
          fill=NA) +
  scale_color_manual(values = c( "red", "cyan", "black"),
                       name = element_blank())+
  guides(fill = guide_legend(order = 1), colour = guide_legend(order = 2))+
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)
```
```{r, eval=FALSE}
SA_im_v_zn208_plot
ggsave("images/SA_im_vv_zn208_plot.png")
```
```{r, eval=FALSE}
all_modes_v_im <- merge(x=car, y=bike, by = "VORI_CD_ZT208", all = T) %>% 
  merge(transit, by = "VORI_CD_ZT208", all=T)%>%
  merge(walk, by = "VORI_CD_ZT208", all = T)

@@

all_modes_v_im_plot <- ggplot(all_modes_v_im,
       aes(x=v_im_transit,
           y = v_im_walk)) +
  geom_line()  + geom_point()+
  theme_classic()
```

# Abstract
An increasing number of studies within the domain of transport are concerned with the inequities in accessibility to opportunities. A dimension of these inequities arise from differences in access by mode type (e.g., the number of work opportunities that can be reached using a car as opposed to transit in a city). However, methods assessing multimodal accessibility in the literature fall short as aspects of competition for opportunities and the explicit methodological acknowledgement of opportunities being _finite_ are lacking. In this vein, this paper presents an extension of _spatial availability_, a singly-constrained competitive accessibility measure, for the context of multimodal accessibility analysis. We first illustrate the features of spatial availability that lends itself to multimodal analysis. We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes. In summary, spatial availability can be used to create and interpret multimodal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape by better identifying differences in accessibility afforded by different modes.

@@

From the perspective of urban transport systems, location-based accessibility measures have been used in the context of policy evaluation. For instance, @leeMeasuringImpactsNew2018 assesses the transit accessibility gains to healthcare and employment opportunities for disadvantaged neighbourhood in Columbus, Ohio, USA after the transit system's re-design and introduction of a rapid bus system. However, a limitation of this study, like others that implement accessibility measures, is they do not calculate results under a _constrained_ framework i.e., one of _finity_. The citizens of Columbus should experience qualitative accessibility gains - but is it at the expense of access to opportunities to populations using other modes? <!-- As another example, @mohriClusteringMethodMeasuring2021 implements a modified cumulative opportunity measure to assess differences between private vehicle and transit system accessibility to jobs in Melbourne, but a similar question remains, does the accessibility afforded to the private vehicle using population come at the expense of accessibility losses to transit users? -->

The work of @leeMeasuringImpactsNew2018 uses _non-competitive_ accessibility measures. There is a branch of location-based accessibility measures that do incorporate the effect of competition for opportunities by the population in the region. However, we argue that these existing methods fall short in acknowledging the _finity_ of opportunities. For instance, @maoMeasuringSpatialAccessibility2013 applies a competitive measure, two-step floating catchment approach (2SFCA), for the case of access to healthcare services in Florida for both a multimodal network and a single modal network. While the differences in modal access are discussed, the question of how the advantage in access afforded by one mode over another impacts access for different mode users is unanswered.

This question of how much one mode-using population can access at the expense of another mode-using population is a pertinent equity question in the evaluation of policy scenarios that are multimodal. For instance, consider the impact of a low emission zone (LEZ). LEZ is a policy of spatial and modal discrimination: the circulation of vehicles that are excessively polluting are restricted in specific regions. In the recognition that opportunities are finite, the implementation of a LEZ explicitly reduces the access that the population using polluting vehicles has to opportunities. This restriction allows the population using other more sustainable modes to potentially have a higher level of access than before the LEZ implementation. This evaluation is especially urgent as LEZ are currently in effect in cities globally; their reception has been mixed [@tarrinoortizPublicAcceptabilityLow2021] and may be having negative impacts on disadvantaged populations who have become mobility restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022]. Measures that evaluate the accessibility of modes given both _constrained_ and _competitive_ considerations are lacking in the literature, but are needed, to evaluate such policy interventions impact on accessibility.

@@

# A review of multimodal accessibility measures
Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are a product of the relationship between land-use and transport systems. Arguably the most commonly used location-based measured are cumulative opportunity measures and weighted cumulative opportunity measures [@levinsonTransportAccessManual2020]. These measures weight the opportunities that can be potentially interacted with from origin $i$ to destination $j$ based on some sort of travel cost function (e.g., travel time, fare, travel distance) otherwise known as a travel impedance function $f^{m}(c^m_{ij})$. Many weighted cumulative opportunities (often refered to as the gravity-based measure) originate from the measure proposed by @hansenHowAccessibilityShapes1959, which can take the following multimodal form: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs $(c_{ij}^m)$ and travel impedance functions $f^m(\cdot)$.

@@

In brief, we define the _spatial availability_ at $i$ ( $V_{i}$ ) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$. $V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region. The general formulation of spatial availability $V_{i}$ is shown in Equation (\ref{eq:spatial-availability-general}):
```{=tex}
\begin{equation}
\label{eq:spatial-availability-general}
V_i = \sum_{j=1}^J O_jF^t_{ij}
\end{equation}
```

@@

Spatial availability measure is introduced in @soukhovIntroducingSpatialAvailability2023. The unique feature in the measure is the balancing factor $F^t_{ij}$, a proportional allocation mechanisms, that ensures $V_i$ calculated for each $i$ sums to the total number of opportunities. Through $F^t_{ij}$, spatial availability is a  _competitive_ and _constrained_ accessibility measure that handles the number of opportunities in the region in a finite way. $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i} = \frac{P_i}{\sum_i P_i}$ and an impedance-based balancing factor $F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population at $i$ (the mass effect) and the cost of reaching opportunities at $j$ (the impedance effect).

@@

<!-- THIS IS GOOD, but cutting to save words for now.

>>>>

---
title: "Multimodal Spatial Availability: A Singly-Constrained Measure of Competitive Accessibility by Multiple Modes"
author:
  - name: Anastasia Soukhov
    note: Corresponding Author
    position: PhD Candidate
    email: soukhoa@mcmaster.ca
    affiliation: Department of Earth, Environment and Society, McMaster University, Canada
  - name: Javier Tarriño-Ortiz
    position: PhD Graduate
    email: javier.tarrino.ortiz@upm.es
    affiliation: Universidad Politécnica de Madrid, Spain
  - name: Julio A. Soria-Lara
    position: Associate Professor
    email: julio.soria-lara@upm.es
    affiliation: Urban and Regional Planning, Universidad Politécnica de Madrid, Spain
  - name: Antonio Páez
    position: Associate Professor
    email: paezha@mcmaster.ca
    affiliation: Department of Earth, Environment and Society, McMaster University, Canada
abstract: |
  An increasing number of studies within the domain of transportation planning are concerned with the inequities in accessibility to opportunities. A dimension of these inequities arises from differences in access by mode type (e.g., commuting using a car as opposed to transit). However, methods implemented in current accessibility literature are lacking within the context of multimodal analysis. This paper presents an extension of spatial availability, a singly-constrained competitive accessibility measure, for the context of multimodal accessibility analysis. We first illustrate the features of spatial availability that lend itself to multimodal analysis. We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes. In summary, spatial availability can be used to create and interpret multimodal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape.

keywords: ["Multimodal", "Accessibility", "Equity", "Policy scenarios", "Low emission zones"]

bibliography: mybibfile.bib
output: rticles::plos_article
# plos.csl can be download and used locally
csl: http://www.zotero.org/styles/plos
---
```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = '', 
  out.width = "1\\linewidth")

@@

library(patchwork) # The Composer of Plots
library(sf) # Simple Features for R
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggforce) # Accelerating 'ggplot2'
library(geomtextpath) # Curved Text in 'ggplot2'
library(sfheaders)
library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
library(ggspatial) # Spatial Data Framework for ggplot2
library(flextable) # Functions for Tabular Reporting
library(scales) # Scale Functions for Visualization
library(TAM) #Test Analysis Modules
library(officer)
library(ggnewscale)
library(cowplot)
```

@@

# od_lines coded as segments
od_lines <- data.frame(x = c(2.5, 2.5, 2.5, 2.5, 2.5, 2.5, 6.5, 6.5, 6.5),
                       y = c(8.0, 8.0, 8.0, 2.5, 2.5, 2.5, 4.5, 4.5, 4.5),
                       xend = c(0.8, 0.4, 6.3, 0.4, 0.8, 6.3, 0.8, 0.7, 6.5),
                       yend = c(7.7, 2.8, 2.7, 7.2, 2.5, 2.5, 7.5, 2.7, 2.7),
                       group = c("A", "A", "A", "B", "B", "B", "C", "C", "C"),
                       time = c("15 min (x)\n10 min (z)", "30 min (x)\n25 min (z)", "100 min (x)\n80 min (z)",
                                "30 min (x)\n25 min (z)", "15 min (x)\n10 min (z)", "100 min (x)\n80 min (z)",
                                "100 min (x)\n80 min (z)", "100 min (x)\n80 min (z)", "15 min (x)\n10 min (z)"))

@@

```{r, creating-toy-example-plot, eval=FALSE}
toy_example_plot <- ggplot() + 
  # Plot centers
  geom_circle(data = centers,
             aes(x0 = x, 
                 y0 = y,
                 r = radius),
             color = "grey40") +
  annotate(geom = "label", 
           x=c(0.5, 0.6, 7),
           y=c(9, 1, 5.2), 
           label = c("Suburban", "Urban", "Satellite"), 
           size=4, 
           label.size=NA) +
  # Connect origins and destinations with curves
  geom_curve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group),
             arrow = arrow(length = unit(0.03, "npc")),
             curvature = 0.20,
             linewidth = 1,
             color = "black") +
  # Text on curves
  geom_textcurve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group,
                   label = time),
              size = 3,
             curvature = 0.20,
             color = "black") +
  # Plot origins and destinations
  geom_point(data = od,
            aes(size = size,
                x = x, 
                y = y, 
                color = type,
                shape = type)) +
  scale_size(range = c(4, 10)) +
  # Label origins and destinations
  geom_text(data = od, 
            aes(x, 
                y, 
                label=id), 
            size=3) +
  # Create table
  annotate(geom = "table",
           x = 8, 
           y = 9, 
           label = list(od_table),
           fill = "white",
           size=3) +
  # Theme the plot
  coord_fixed() + 
  theme_void() + guides(shape = "none", size = "none", color = "none", linetype ="none")

@@

sf::sf_use_s2(FALSE) #this terns off the spherical geometry component in the st_union() function - this assumes the sf is just planar. Thats' fine for our purposes
Madrid_muni_border <- zn1259_mad %>%
  st_union()
```
```{r merging-borders-for-plotting, warning=FALSE}
Madrid_muni_border <- Madrid_muni_border %>% st_as_sf() %>% transmute(Border = "Madrid municipal border") %>% rename( "geometry" = "x")

borders <- rbind(LEZ_centro %>% transmute(Border = "LEZ centro"), 
                 LEZ_M30 %>% transmute(Border = "LEZ M-30"),
                 Madrid_muni_border)

#to clean up the slivers
borders <- borders |> st_make_valid() |>
  group_by(Border) |>
  summarise() |>
  sf_remove_holes()
```

@@

```{r}
#Plot v (spatial availability per capita) by origin:
SA_im_v_zn208_plot <- ggplot() +
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=NA,
          fill="grey30") +
  geom_sf(data = SA_im_zn208,
          aes(fill = v_im),
          color = NA) +
  scale_fill_gradient2(low="red", high="green", mid = "white", midpoint = 1,
                       name = expression(v["i"]^{m}),
                       limits = c(0.001106867, 2.086736720),
                       breaks = c( 0.12, 0.31, 0.80, 1.11, 2.086736720),
                       labels = c(">0.12 (20% prob.)", "0.31 (40% prob.)", "0.80 (60% prob.)", "1.11 (80% prob.)", "2.09 (max.)"))+
    geom_sf(data = borders,
          aes(colour=Border),
          size = 0.5,
          fill=NA) +
  scale_color_manual(values = c( "red", "cyan", "black"),
                       name = element_blank())+
  guides(fill = guide_legend(order = 1), colour = guide_legend(order = 2))+
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)
```
```{r, eval=FALSE}
SA_im_v_zn208_plot
ggsave("images/SA_im_vv_zn208_plot.png")
```

all_modes_v_im_plot <- ggplot(all_modes_v_im,
       aes(x=v_im_transit,
           y = v_im_walk)) +
  geom_line()  + geom_point()+
  theme_classic()
```
```{r, eval=FALSE}
all_modes_v_im <- merge(x=car, y=bike, by = "VORI_CD_ZT208", all = T) %>% 
  merge(transit, by = "VORI_CD_ZT208", all=T)%>%
  merge(walk, by = "VORI_CD_ZT208", all = T)

@@

```{r, eval=FALSE}
#Check that the totals are maintained, it is!:
sum(OD_SA_zn208$V_ij, na.rm = T) #847574 jobs spatially avaialble to the region -- 
OD_SA_zn208 %>% group_by(VDES_CD_ZT208) %>% summarize(jobs = mean(jobs)) %>% select(jobs) %>% sum()

#for comparision, the number of travel minutes spent going to jobs. group by jobs zona, and the the number of potential minutes traveling to jobs. This sum is the number of minutes spent commuting to jobs. it is 33680009 minutes, this is approx equal to 43min (mean travel time) *847574 (the total number of workers).
OD_SA_zn208 %>% group_by(VDES_CD_ZT208) %>% 
  summarize(jobs_tt_f_t = sum(jobs_tt_f_t)) %>% 
  select(jobs_tt_f_t) %>% sum() #should equal 33680009 mins total
```

```{r}
#Maybe delete this chunk??

# Spatial availability by mode, we can see that car mode captures ~48% of the spatial availability but only accounts for ~36% of the tt_ijm !!:
modal_V_comps <- OD_SA_zn208 %>%
  group_by(mode) %>%
  summarize(workers_m = sum(workers),
            V_m = sum(V_ij),
            V_m_perc = sum(V_ij)/847574,
            jobs_tt_f_t = sum(jobs_tt_f_t)/33680009)
#modal_V_comps #this issue here is workers should equal V_m in our case, make it do it.
```

```{r}
#data for the following plot
modal_V_comps_4plot <- SA_im_zn208 %>% st_drop_geometry() %>%
  group_by(mode) %>% 
  summarize(workers_m = sum(population_m),
            v_m = mean(v_im),
            V_m = sum(V_im) ,
            V_m_perc = sum(V_im)/847574,
            workers_m_perc = sum(population_m)/847574)

modal_V_comps_4plot <- data.frame(Mode = c("Bike", "Walk", "Car/motor", "Transit",
                                           "Bike", "Walk", "Car/motor", "Transit"),
                                  perc = c(modal_V_comps_4plot$workers_m_perc, modal_V_comps_4plot$V_m_perc),
                                  Measure = c(1,1,1,1,2,2,2,2) ) #for labeling the y-axis

CENTRO_comps <- SA_im_zn208 %>% st_drop_geometry() %>% filter(VORI_CD_ZT208 <= 3) %>%
  group_by(mode) %>% 
  summarize(workers_m = sum(population_m),
            V_m = sum(V_im)) %>%
  transmute(mode,
    CENTRO_V_m_perc = V_m/sum(V_m),
    CENTRO_workers_m_perc = workers_m/sum(workers_m))

CENTRO_comps <- data.frame(Mode = c("Bike", "Walk", "Car/motor", "Transit",
                                           "Bike", "Walk", "Car/motor", "Transit"),
                                  perc = c(CENTRO_comps$CENTRO_workers_m_perc, CENTRO_comps$CENTRO_V_m_perc),
                                  Measure = c(5,5,5,5,6,6,6,6) ) #for labeling the y-axis

M30_comps <- SA_im_zn208 %>% st_drop_geometry() %>% filter(VORI_CD_ZT208 >= 4 & VORI_CD_ZT208 <= 20) %>%
  group_by(mode) %>% 
  summarize(workers_m = sum(population_m),
            V_m = sum(V_im)) %>%
  transmute(mode,
    M30_V_m_perc = V_m/sum(V_m),
    M30_workers_m_perc = workers_m/sum(workers_m))

M30_comps <- data.frame(Mode = c("Bike", "Walk", "Car/motor", "Transit",
                                           "Bike", "Walk", "Car/motor", "Transit"),
                                  perc = c(M30_comps$M30_workers_m_perc, M30_comps$M30_V_m_perc),
                                  Measure = c(3,3,3,3,4,4,4,4) )

modal_V_comps_4plot <- rbind(modal_V_comps_4plot,M30_comps,CENTRO_comps)
```

```{r}
#comparision plots of spatial availability vs. population by mode for the full region, just m-30, and just centro
comps_plot <- ggplot(data = modal_V_comps_4plot,
       aes(y=perc, x=Measure, fill=Mode, label= perc %>% percent(accuracy = 0.1) )) +
  scale_x_discrete(limits = c("Population by Mode\n(Total: 847,574)","Spatial Availability\n(Total: 847,574)",
                              "Population by Mode\n(Total: 243,472)","Spatial Availability\n(Total: 217,219)",
                              "Population by Mode\n(Total: 44,891)","Spatial Availability\n(Total: 38,942)"),
                   position = "bottom") +
  geom_col() +
  scale_fill_manual(values = c("lightgreen","sienna", "coral", "skyblue"))+
  geom_text(position = position_stack(vjust=0.5), color="black")+
  theme_classic()+
  theme(axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.line = element_blank(),
        axis.text.y = element_blank(),
        axis.text = element_text(color="black"),
        legend.position = "bottom",
        legend.title = element_blank()) +
  geom_vline(xintercept=c(0.5,2.5, 4.5, 6.5),
                color = "black", size=1)
```

```{r}
ALL_Plot <- ggplot() +
    geom_sf(data = borders,
          size = 0.5,
          aes(col = Border),
          fill= NA) +
  scale_color_manual(values = c( "red", "cyan", "black"),
                     name = element_blank())+
  geom_sf(data = borders %>% filter(Border == "Madrid municipal border"),
          size = 0.5,
          col ="black",
          fill="black") +
  theme_void() +
  ggtitle("All") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")

M30_Plot <- ggplot() +
    geom_sf(data = borders,
          size = 0.5,
          aes(col = Border),
          fill= NA) +
  scale_color_manual(values = c( "red", "cyan", "black"),
                     name = element_blank())+
  geom_sf(data = borders %>% filter(Border == "LEZ M-30"),
          size = 0.5,
          col ="cyan",
          fill="cyan") +
  theme_void() +
  ggtitle("M-30") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")

CENTRO_Plot <- ggplot() +
    geom_sf(data = borders,
          size = 0.5,
          aes(col = Border),
          fill= NA) +
  scale_color_manual(values = c( "red", "cyan", "black"),
                     name = element_blank())+
  geom_sf(data = borders %>% filter(Border == "LEZ centro"),
          size = 0.5,
          col ="red",
          fill="red") +
  theme_void() +
  ggtitle("Centro") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")

modal_V_comps_4plot_plot <- gridExtra::arrangeGrob(
                        #labels_4plot,
  gridExtra::arrangeGrob(ALL_Plot, M30_Plot, CENTRO_Plot, ncol=3),
                        comps_plot,
                        nrow=2, heights = c(1,2.5) )
ggsave("images/modal_V_comps_4plot.png", modal_V_comps_4plot_plot)
```

# Abstract
An increasing number of studies within the domain of transport are concerned with the inequities in accessibility to opportunities. A dimension of these inequities arise from differences in access by mode type (e.g., the number of work opportunities that can be reached using a car as opposed to transit in a city). However, methods assessing multimodal accessibility in the literature fall short as aspects of competition for opportunities and the explicit methodological acknowledgement of opportunities being _finite_ are lacking. In this vein, this paper presents an extension of _spatial availability_, a singly-constrained competitive accessibility measure, for the context of multimodal accessibility analysis. We first illustrate the features of spatial availability that lends itself to multimodal analysis. We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility for locations where populations use different modes. In summary, spatial availability can be used to create and interpret multimodal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape by better identifying differences in accessibility afforded by different modes.

@@

From the perspective of urban transport systems, location-based accessibility measures have been used in the context of policy evaluation. For instance, @leeMeasuringImpactsNew2018 assesses the transit accessibility gains to healthcare and employment opportunities for disadvantaged neighbourhood in Columbus, Ohio, USA after the transit system's re-design and introduction of a rapid bus system. However, a limitation of this study, like others that implement accessibility measures, is they do not calculate results under a _constrained_ framework i.e., one of _finity_. The citizens of Columbus should experience quantitative accessibility gains - but is it at the expense of access using other modes? As another example, @mohriClusteringMethodMeasuring2021 implements a modified cumulative opportunity measure to assess differences between private vehicle and transit system accessibility to jobs in Melbourne, but a similar question remains: does the accessibility afforded to the private vehicle using population come at the expense of accessibility losses to transit users?

@leeMeasuringImpactsNew2018 and @mohriClusteringMethodMeasuring2021 both use _non-competitive_ accessibility measures. There is a branch of location-based accessibility measures that do incorporate the effect of competition for opportunities by the population in the region. However, we argue that these existing methods fall short in acknowledging the _finity_ of opportunities. For instance, @maoMeasuringSpatialAccessibility2013 applies a competitive measure, two-step floating catchment approach (2SFCA), for the case of access to healthcare services in Florida for both a multimodal network and a single modal network. While the differences in modal access are discussed, the question of how the advantage in access afforded by one mode over another impacts access for different mode users is unanswered.

This question of how much one mode-using population can access at the expense of another mode-using population is a pertinent equity question in the evaluation of policy scenarios that are multimodal. For instance, consider the impact of a low emission zone (LEZ). LEZ is a policy of spatial and modal discrimination: the circulation of vehicles that are excessively polluting are restricted in specific regions. In the recognition that opportunities are finite, the implementation of a LEZ explicitly reduces the access that the population using polluting vehicles has to opportunities. This restriction allows the population using other more sustainable modes to potentially have a higher level of access than before the LEZ implementation. This evaluation is especially urgent as LEZ are currently in effect in cities globally; their reception has been mixed [@tarrinoortizPublicAcceptabilityLow2021] and may be having negative impacts on disadvantaged populations who have become mobility-restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022]. Measures that evaluate the accessibility of modes given both _constrained_ and _competitive_ considerations are lacking in the literature, but are needed, to evaluate such policy interventions impact on accessibility.

@@

# A review of multimodal accessibility measures
Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are a product of the relationship between land-use and transport systems. Arguably the most commonly used location-based measured are cumulative opportunity measures and weighted cumulative opportunity measures [@levinsonTransportAccessManual2020]. These measures weight the opportunities that can be potentially interacted with from origin $i$ to destination $j$ based on some sort of travel cost function (e.g., travel time, fare, travel distance) otherwise known as a travel impedance function $f^{m}(c^m_{ij})$. Many weighted cumulative opportunities (often referred to as the gravity-based measure) originate from the measure proposed by @hansenHowAccessibilityShapes1959, which can take the following multimodal form: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs $(c_{ij}^m)$ and travel impedance functions $f^m(\cdot)$.

@@

In brief, we define the _spatial availability_ at $i$ ($V_{i}$) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$. $V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region. The general formulation of spatial availability $V_{i}$ is shown in Equation (\ref{eq:spatial-availability-general}):
```{=tex}
\begin{equation}
\label{eq:spatial-availability-general}
V_i = \sum_{j=1}^J O_jF^t_{ij}
\end{equation}
```

@@

The spatial availability measure is introduced in @soukhovIntroducingSpatialAvailability2023. Spatial availability's unique feature is the balancing factor $F^t_{ij}$, a proportional allocation mechanisms, that ensures the $V_i$ calculated for each $i$ sums, across all $i$ in the region, to equal the total number of opportunities in the region. As such, spatial availability is a  _competitive_ and _constrained_ accessibility measure as $F^t_{ij}$ handles the number of opportunities in the region in a finite way (proportional allocation). $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i} = \frac{P_i}{\sum_i P_i}$ and an impedance-based balancing factor $F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population at $i$ (the mass effect) and the cost of reaching opportunities at $j$ (the impedance effect).

@@

<!-- THIS IS GOOD, but could be cut to save works-->
Since spatial availability is constrained and has an interpretable meaning as a proportion of the total opportunities in the region, the values at $i$ have a new significance. Inequality in $V_i^m$ values can be explored through a variety of approaches. For instance, consider travel times. The $z$-using population accounts for `r ((V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% filter(m == "z") %>% dplyr::select(V_tt) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` of the potential travel time traveled in the region: this is `r (((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) - (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% filter(m == "z") %>% dplyr::select(V_tt) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` less travel time than the proportion of spatial available opportunities that is allocated to them. In other words, the $z$-using population travels less minutes overall and has more spatial availability of opportunities than the $x$-using population using the slower mode $x$.

Alternatively, inequities in spatial availability between mode-using populations can be explored through proportional benchmarks. A spatial availability per capita $v_i^m$ as presented in Equation (\ref{eq:SA-per-capita}):
